Magnetised Polish doughnuts revisited

### 2017-04-04 paper.tex
final version including Fragile and Stergioulas comments

  \abstract
{We discuss a procedure to build new sequences of magnetised, equilibrium tori around Kerr black holes which combines two approaches previously considered in the literature. For simplicity we assume that the test-fluid approximation holds, and hence we neglect the self-gravity of the fluid. The models are built assuming a particular form of the angular momentum distribution from which the location and morphology of surfaces of equal pressure can be computed. This ansatz includes, in particular, the  constant angular momentum case originally employed in the construction of thick tori -- or Polish doughnuts -- and it has already been used to build equilibrium sequences of purely hydrodynamical models. We discuss the properties of the new models and their dependence on the initial parameters. These new sequences can be used as initial data for magnetohydrodynamical evolutions in general relativity.}

@@

 In this paper we combine the two approaches considered in~\cite{Komissarov:2006} and~\citet{Qian:2009} to build new sequences of equilibrium tori around Kerr black holes. Building on these works, we present here the extension of the models of~\citet{Qian:2009} to account for discs endowed with a purely toroidal magnetic field. In our procedure we hence assume a form of the angular momentum distribution that departs from the constant case considered by~\cite{Komissarov:2006} and from which the location and morphology
of surfaces of equal pressure can be numerically computed. As we shall show below, for the particular case of constant angular momentum distributions, our method is in good agreement with the results of~\citet{Komissarov:2006}.

@@

To reduce the number of free parameters to build the initial models we do as in~\citet{Komissarov:2006} and fix the values of the equation of state exponents, $\kappa$ and $\lambda$, and of the enthalpy at the disc centre, $w_{\mathrm{c}}$. More precisely, we choose $\kappa = \lambda = 4/3$ and $w_{\mathrm{c}} = 1$. This still leaves us with five parameters to control the size, shape, thickness, and magnetisation of the disc, namely the magnetisation parameter $\beta_{\mathrm{m}_{\mathrm{c}}}$, the parameters of the angular momentum distribution $\beta$ and $\gamma$, the black hole spin parameter $a$, and the inner radius of the disc $r_{\mathrm{in}}$. We build a series of 45 models, whose main features are summarised in Table~\ref{table:1} (for $\beta_{\mathrm{m}_{\mathrm{c}}} = 10^{3}$, i.e.~models where the effects of the magnetisation are unimportant), Table~\ref{table:2} ($\beta_{\mathrm{m}_{\mathrm{c}}} = 1$, i.e.~equipartition models), and Table~\ref{table:3} ($\beta_{\mathrm{m}_{\mathrm{c}}} = 10^{3}$, i.e.~highly-magnetised models).

@@

A representative sample of the isodensity surfaces for some of our models appears in Figure~\ref{models}. We plot, in particular, the models of Table~\ref{table:3}, for which the value of the magnetisation parameter is the highest, $\beta_{\mathrm{m}_{\mathrm{c}}}=10^{-3}$. From left to right the columns of this figure correspond to increasing values of the black hole spin, namely, $a=0.5$, 0.9 and 0.99, while from top to bottom the rows correspond to different combinations of the $\gamma$ and $\beta$ parameters that characterize the ansatz for the angular momentum distribution of~\citet{Qian:2009} (the particular values are indicated in the caption of the figure).  A rapid inspection shows that the structure of the discs noticeable changes when the parameters change. Notice that the spatial scale in all of the plots of this figure has been chosen so as to facilitate the visualization of the discs (which can be fairly small in some cases) and, as such, is different in each plot. A typical Polish doughnut is represented by the model in the top-left panel. As the black hole spin increases, the discs are closer to the black hole and its relative size with respect to the black hole is smaller for all values of $\gamma$ and $\beta$. For $\gamma=0$ the discs are infinite, irrespective of the value of $a$ and $\beta$. As $\gamma$ and $\beta$ increase from 0 to 0.9 (compare the three top rows) the discs become significantly thinner and radially smaller (see also Fig.~\ref{more-profile} below). It is also interesting to note that the maximum value of the density is higher (with respect to the value at the disc centre) as the spin increases.

@@

In this paper we have presented a procedure to build equilibrium sequences of magnetised, non-self-gravitating discs around Kerr black holes which combines the two existing approaches of~\citet{Komissarov:2006} and~\citet{Qian:2009}. On the one hand we have followed~\citet{Qian:2009} and have assumed a form of the angular momentum distribution in the disc from which the location and morphology of surfaces of equal pressure can be computed. As a limiting case, this ansatz includes the constant angular momentum case originally employed in the construction of thick tori -- or Polish doughnuts~\citep{Abramowicz:1978,Kozlowski:1978}  -- and was already used by~\citet{Qian:2009} to build equilibrium sequences of purely hydrodynamical models. On the other hand, our discs are endowed with a purely toroidal magnetic field, as in the work of~\citet{Komissarov:2006}, which provides the methodology we have followed to handle the magnetic terms.

@@

\begin{acknowledgements}
      Work supported by the Spanish MINECO (grant AYA2015-66899-C2-1-P) and the Generalitat Valenciana (PROMETEOII-2014-069).
\end{acknowledgements}

>>>>

  \abstract
{We discuss a procedure to build new sequences of magnetised, equilibrium tori around Kerr black holes which combines two approaches previously considered in the literature. For simplicity we assume that the test-fluid approximation holds, and hence we neglect the self-gravity of the fluid. The models are built assuming a particular form of the angular momentum distribution from which the location and morphology of equipotential surfaces can be computed. This ansatz includes, in particular, the  constant angular momentum case originally employed in the construction of thick tori -- or Polish doughnuts -- and it has already been used to build equilibrium sequences of purely hydrodynamical models. We discuss the properties of the new models and their dependence on the initial parameters. These new sequences can be used as initial data for magnetohydrodynamical evolutions in general relativity.}

@@

 In this paper we combine the two approaches considered in~\cite{Komissarov:2006} and~\citet{Qian:2009} to build new sequences of equilibrium tori around Kerr black holes. Building on these works, we present here the extension of the models of~\citet{Qian:2009} to account for discs endowed with a purely toroidal magnetic field. In our procedure we hence assume a form of the angular momentum distribution that departs from the constant case considered by~\cite{Komissarov:2006} and from which the location and morphology of the equipotential surfaces can be numerically computed. As we shall show below, for the particular case of constant angular momentum distributions, our method is in good agreement with the results of~\citet{Komissarov:2006}. We also note the recent work of~\citet{Wielgus:2015} where Komissarov's solution was extended for the particular case of power-law distributions of angular momentum. Moreover, the magnetised tori of~\citet{Wielgus:2015} were used to explore the growth of the magneto-rotational instabilty (MRI) through time-dependent numerical simulations. In particular, the long-term evolution of those tori has been recently investigated by~\citet{Fragile:2017}, who paid special attention to the decay of their magnetisation.

@@

To reduce the number of free parameters to build the initial models we do as in~\citet{Komissarov:2006} and fix the values of the equation of state exponents, $\kappa$ and $\lambda$, and of the enthalpy at the disc centre, $w_{\mathrm{c}}$. More precisely, we choose $\kappa = \lambda = 4/3$ and $w_{\mathrm{c}} = 1$. This still leaves us with five parameters to control the size, shape, thickness, and magnetisation of the disc, namely the magnetisation parameter $\beta_{\mathrm{m}_{\mathrm{c}}}$, the parameters of the angular momentum distribution $\beta$ and $\gamma$, the black hole spin parameter $a$, and the inner radius of the disc $r_{\mathrm{in}}$. We build a series of 45 models, whose main features are summarised in Table~\ref{table:1} (for $\beta_{\mathrm{m}_{\mathrm{c}}} = 10^{3}$, i.e.~models where the effects of the magnetisation are unimportant), Table~\ref{table:2} ($\beta_{\mathrm{m}_{\mathrm{c}}} = 1$, i.e.~equipartition models), and Table~\ref{table:3} ($\beta_{\mathrm{m}_{\mathrm{c}}} = 10^{-3}$, i.e.~highly-magnetised models).

@@

A representative sample of the isodensity surfaces for some of our models appears in Figure~\ref{models}. We plot, in particular, the models of Table~\ref{table:3}, for which the magnetisation is highest ($\beta_{\mathrm{m}_{\mathrm{c}}}=10^{-3}$). From left to right the columns of this figure correspond to increasing values of the black hole spin, namely, $a=0.5$, 0.9 and 0.99, while from top to bottom the rows correspond to different combinations of the $\gamma$ and $\beta$ parameters that characterize the ansatz for the angular momentum distribution of~\citet{Qian:2009} (the particular values are indicated in the caption of the figure).  A rapid inspection shows that the structure of the discs noticeable changes when the parameters change. Notice that the spatial scale in all of the plots of this figure has been chosen so as to facilitate the visualization of the discs (which can be fairly small in some cases) and, as such, is different in each plot. A typical Polish doughnut is represented by the model in the top-left panel. As the black hole spin increases, the discs are closer to the black hole and its relative size with respect to the black hole is smaller for all values of $\gamma$ and $\beta$. For $\gamma=0$ the discs are infinite, irrespective of the value of $a$ and $\beta$. As $\gamma$ and $\beta$ increase from 0 to 0.9 (compare the three top rows) the discs become significantly thinner and radially smaller (see also Fig.~\ref{more-profile} below). It is also interesting to note that the maximum value of the density is higher (with respect to the value at the disc centre) as the spin increases.

@@

In this paper we have presented a procedure to build equilibrium sequences of magnetised, non-self-gravitating discs around Kerr black holes which combines the two existing approaches of~\citet{Komissarov:2006} and~\citet{Qian:2009}. On the one hand we have followed~\citet{Qian:2009} and have assumed a form of the angular momentum distribution in the disc from which the location and morphology of surfaces of equal potential can be computed. As a limiting case, this ansatz includes the constant angular momentum case originally employed in the construction of thick tori -- or Polish doughnuts~\citep{Abramowicz:1978,Kozlowski:1978}  -- and was already used by~\citet{Qian:2009} to build equilibrium sequences of purely hydrodynamical models. On the other hand, our discs are endowed with a purely toroidal magnetic field, as in the work of~\citet{Komissarov:2006}, which provides the methodology we have followed to handle the magnetic terms. On a similar note, we cite the work of~\citet{Wielgus:2015} who have recently extended the solution of~\citet{Komissarov:2006} to include non-constant specific angular momentum tori. These authors  limited their consideration to power-law distributions and were particularly focused on the stability of such tori to the MRI. The approach discussed in our work differs from that of~\citet{Wielgus:2015} and, moreover, we have explored a much wider parameter space.

@@

\begin{acknowledgements}
It is a pleasure to thank P.~C.~Fragile and N.~Stergioulas for useful discussions and comments. Work supported by the Spanish MINECO (grant AYA2015-66899-C2-1-P) and the Generalitat Valenciana (PROMETEOII-2014-069).
\end{acknowledgements}
